%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,review, anonymous]{acmart}


%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
    
% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{2018}
\acmYear{2018}
\setcopyright{acmlicensed}
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmDOI{10.1145/1122445.1122456}
\acmISBN{978-1-4503-9999-9/18/06}

%
% These commands are for a JOURNAL article.
%\setcopyright{acmcopyright}
%\acmJournal{TOG}
%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
%\acmDOI{10.1145/1122445.1122456}

%
% Submission ID. 
% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
% of the event, and this ID should be used as the parameter to this command.
%\acmSubmissionID{123-A56-BU3}

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
%\citestyle{acmauthoryear}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}


%% Tikz
\usetikzlibrary{positioning}
\tikzset{main node/.style={circle,fill=gray!20,draw,minimum size=0.45cm,inner sep=0pt}, }
\tikzset{invisible/.style={circle,fill=white!20,draw,minimum size=0.0cm,inner sep=0pt}, }


%
% end of the preamble, start of the body of the document source.
\begin{document}

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Distributed Relational Algebra at Scale}

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.
\author{Sidharth Kumar}
\authornote{Both authors contributed equally to this research.}
\email{sid14@uab.edu}
\orcid{}
\author{Thomas Gilray}
\authornotemark[1]
\email{gilray@uab.edu}
\affiliation{%
  \institution{University of Alabama, Birmingham}
  \streetaddress{}
  \city{Birmingham}
  \state{Alabama}
  \postcode{35233}
}

\newenvironment{tightenumerate}{
	\edef\backupindent{\the\parindent}
	\begin{enumerate}
		\setlength{\itemsep}{1pt}
		\setlength{\parskip}{0pt}
		\setlength{\parsep}{0pt}
		\setlength{\parindent}{\backupindent}
	}{\end{enumerate}
}
\newenvironment{tightitemize}{
	\edef\backupindent{\the\parindent}
	\begin{itemize}
		\setlength{\itemsep}{1pt}
		\setlength{\parskip}{0pt}
		\setlength{\parsep}{0pt}
		\setlength{\parindent}{\backupindent}
	}{\end{itemize}
}

%\author{Julius P. Kumquat}
%\affiliation{\institution{The Kumquat Consortium}}
%\email{jpkumquat@consortium.net}

%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.
\renewcommand{\shortauthors}{Kumar and Gilray}

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
%Relational algebra forms a basis of primitive operations useful for applications in graphs, networks, program analysis, deductive databases, and logic. Despite its expressive power, relational algebra has not received the same attention in high-performance computing research as linear algebra, stencil computation, map-reduce and graph analytics.

%In this paper we present a set of efficient algorithms that tackle the problem of distributed, parallel relational algebra and use experiments from applications in graphs, program analysis, and datalog to evaluate our approach.
  Relational algebra forms a basis of primitive operations suitable for applications in graphs and networks, program analysis, deductive databases, and constraint logic programming. Despite its expressive power, relational algebra has not received the same attention in high-performance-computing research as more common primitives like stencil computations, floating-point operations, numerical integration, and sparse linear algebra. Furthermore, specific challenges in addressing representation and communication among distributed portions of a relation have previously thwarted successful scaling of relational algebra applications to supercomputers.
  
In this paper, we present a set of efficient algorithms to effectively parallelize and scale key relational algebra primitives. We introduce a hybrid hash-tree approach to representing distributed relations and permitting efficient communication. Finally, we demonstrate the scalability of our implementation with a fixed-point algorithm computing the transitive closure of a large graph (generating over $276$ billion edges) on $65,\!536$ nodes.
\end{abstract}

%
% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
\keywords{datasets, neural networks, gaze detection, text tagging}



%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.
\maketitle

\input{sec-intro}

\input{sec-ra}

\input{sec-impl}

\input{sec-eval}

\input{sec-related}

\section{Conclusion}

We have presented the first general algorithms for scalable relational algebra on supercomputers. Our approach addresses both representation and communication among portions of a distributed relation, laying the groundwork for scaling algorithms that require a pipeline of repeated operations on relations, or fixed-point iteration, such as logical and constraint problems, deductive databases, and static program analyses. We discovered that MPI's all-to-all communication paradigm is scalable, and suitable for relational algebra. Finally, we demonstrated the scalability of our operations, up to $65,\!536$ nodes, in the context of a fixed-point algorithm for computing transitive closure.

In future work, we plan to address distributing relations via hashing on multiple variables to improve performance for highly non-uniform relations. We also have plans to address combining task-level and data-level parallelism in the context of Datalog solvers---a problem that must naturally extend to large numbers of relations and fixed-point interation over equally large numbers of constraints.




%\begin{acks}
%To Robert, for the bagels and explaining CMYK and color spaces.
%\end{acks}

%

% The next two lines define the bibliography style to be used, and the bibliography file.
%\bibliographystyle{ACM-Reference-Format}
\bibliographystyle{abbrv}
\bibliography{sample-base}

% 
% If your work has an appendix, this is the place to put it.
%\appendix


\end{document}
